Investigating Neolithic pottery usage in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-01-18 analysis/paper/paper.Rmd
worked on the discussion section

<!--Info about Guijiabao-->
Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the basin is an unique site with relatively complete and continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b]. Investigation in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and remains, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long term-occupation [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (5000-3700 BP.), the Bronze Age (3200-2700 BP.), and the late North Song to early Ming Dynasties (1200-500 BP.) [@Hao2022]. For the Neolithic period, it can be further separated into two subphases, 5000-4500 cal. BP and 4500-3700 cal. BP, according to the artifact assemblages and ceramic forms and styles. The archaeobotanical remains found at Guidiabao reveal the evidence of crop cultivation dominated by foxtail and broomcorn millet can be traced back to 5000 B.P and a mixed cultivation of millet and rice in a later period (3000–1700 B.C) [@Guijiabao2014; @Hao2022]

@@

## Samples
We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX). The four ceramic fragments were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season. To understand the content and use of pottery vessels in the Yanyan region in an earlier period, we analyze potsherds from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.) with potsherds from a later period (1000 BP.) for comparison. During the early Neolithic period, the common vessel forms include jars, urns, bowls, and basins. For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022]. Our samples are jar fragments from those two common types that are supposed to be used for food serving or storage [].

@@

# Discussion
The results show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant source or marine source. Our striking finding might suggest that consumption of rice could be part of the daily life at Guijiabao and the adoption of rice agriculture could be as important as millet cultivation which was thought the main crop previously. Our next steps is to increase sample size to include pots from different time periods and different types of pottery to identify any variation through time and also try different methods, such as phytolith analysis to further support our statement about rice cultivation.

>>>>

<!--Info about Guijiabao-->
Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the basin is a unique site with relatively complete and continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b]. Investigation in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long-term occupation [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (5000-3700 BP.), the Bronze Age (3200-2700 BP.), and the late North Song to early Ming Dynasties (1200-500 BP.) [@Hao2022]. For the Neolithic period, it can be further separated into two subphases, 5000-4500 cal. BP and 4500-3700 cal. BP, according to the artifact assemblages and ceramic forms and styles. The archaeobotanical remains found at Guidiabao reveal the evidence of crop cultivation dominated by foxtail and broomcorn millet can be traced back to 5000 B.P and mixed cultivation of millet and rice in a later period (3000–1700 B.C) [@Guijiabao2014; @Hao2022].

<!--Info about Gaoshan-->
Gaoshan site is located in the southwestern part of Chengdu Plain.

@@

## Samples
We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX). The four ceramic fragments were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season. To understand the content and use of pottery vessels in the Yanyan region in an earlier period, we analyze potsherds from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.) with potsherds from a later period (1000 BP.) for comparison. During the early Neolithic period, the common vessel forms include jars, urns, bowls, and basins. For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022]. Our samples are jar fragments from those two common types that are supposed to be used for food serving or storage due to the archaeological context of settlement [@Hao2022; @Guijiabao2017a].

@@

# Discussion

Our preliminary results show that the food categories indicated by potsherds at Guijiabao and Gaushan could be mainly dominated by terrestrial animals, such as ruminant animals. Although the yield of lipids in most of our potsherd samples is generally scarce, one potsherd demonstrates a relatively rich composition of long-chain fatty acids which confirms their usage as a cooking or storage pot. A larger amount of C16 and C18 fatty acids compared to other types of fatty acids indicate that the organic residues in our potsherds were highly degraded over the past 4500 years. It is also striking to find that one potsherd of Guijiabao contains residues that might originate from C3 plants.

show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant source or marine source. Our striking finding might suggest that consumption of rice could be part of the daily life at Guijiabao and the adoption of rice agriculture could be as important as millet cultivation which was thought the main crop previously. Our next steps is to increase sample size to include pots from different time periods and different types of pottery to identify any variation through time and also try different methods, such as phytolith analysis to further support our statement about rice cultivation.
